Mathematical Aspects of the Capital Market

### 2017-08-11 papers/FX/PaperBody.tex
Revised my phd thesis.

\begin{abstract}
%We consider an investor with preferences that accord with Generalized
%Disappointment Aversion. Such an investor cares about downside risk and
%we assume he recognizes the heavy tail feature of asset return
%distributions. We argue that when a market is dominated by rational
%investors of this kind, the return distributions of equities that are
%actively traded in this market must have very similar tail-indices. We
%also show, in contrast, the scale parameters of the return
%distributions may differ hugely from one another.
%On the other hand, whether or not all equities in a multivariate model
%have the same tail-index is a dividing issue for multivariate GARCH
%models proposed in the literature. Therefore, it is important to analyze
%data of real equity returns and see how close to each other the
%tail-indices actually are.
%In this work empirical results are also presented and they appear
%to support the conclusion that the tail-indices are very similar,
%with respect to the confidence bands of estimation.
\end{abstract}

>>>>

\begin{abstract}
We consider an investor with preferences that accord with Generalized
Disappointment Aversion (GDA). Such an investor cares about downside
risk and we assume he recognizes the heavy tail feature of asset return
distributions. We argue that when a market is dominated by rational
investors of this kind, the return distributions of equities that are
actively traded in this market may have nearly equal tail-indices due
to monotonicity of the GDA preference with respect to the tail index.

We estimate tail indices and scale parameters of S\&P 500 stocks and
test the hypothesis that two given stock return series have the same
tail index.
% We also show, in contrast, the scale parameters of the return
%distributions may differ hugely from one another.
%On the other hand, whether or not all equities in a multivariate model
%have the same tail-index is a dividing issue for multivariate GARCH
%models proposed in the literature. Therefore, it is important to analyze
%data of real equity returns and see how close to each other the
%tail-indices actually are.
%In this work empirical results are also presented and they appear
%to support the conclusion that the tail-indices are very similar,
%with respect to the confidence bands of estimation.
\end{abstract}

### 2017-08-11 phd/Chapterintroduction.tex
Revised my phd thesis.

\section{GARCH models}
Introduced by Bollerslev \cite{bollerslev:1986} in 1986,
{\em Generalized Autoregressive Conditional Heteroscedasticity}
(GARCH) models have been hugely popular for modelling volatility of
financial time series and have inspired numerous variants.
A GARCH($p, q$) model is a stationary process
$\{X_t\}_{t \in \mathbb Z}$satisfying
\begin{eqnarray*}
  X_t &=& \sigma_t Z_t \\
  \sigma_t^2 &=& \omega + \sum_{i=1}^p \alpha_i X_{t-i}^2 +
  \sum_{j=1}^q \beta_j \sigma_{t-j}^2
\end{eqnarray*}
where $\{X_t\}$ is a return series, e.g. stock returns, foreign exchange
rates, interest rates, etc; $\{Z_t\}$ is an iid, mean 0, unit-variance
sequence, $\sigma_t^2$ is the variance of the distribution of $X_t$
conditional on
$\{(X_i, \sigma_i^2)\}_{i=0}^{t-1}$; $\omega, \{\alpha_i\}_{i=1}^p,
\{\beta_i\}_{i=1}^q$
are non-negative parameters of the model. The GARCH($p,q$) recurrence
equation is of the form of \eqref{eq:rhjyu}. With appropriate
conditions, $\sigma_t^2$ is shown to be a positive Harris recurrent
Markov chain (cf. Bollerslev \cite{bollerslev:1986} and Buraczewski et al.
\cite{buraczewski:damek:mikosch:2016}), whose stationary distribution
has regularly varying tails. The tail index $\alpha$, is given by
\begin{equation}
  \label{eq:grgbg}
  \lim_{n \to \infty} {1 \over n}\log\E\|A_n \cdots A_1\|^\alpha = 0,
\end{equation}
where $\{A_i\}_{i \in \mathbb Z}$ are iid matrices whose entries are
functions of $\{\alpha_i\}_{i=1}^p$, $\{\beta_i\}_{i=1}^q$ and
$\{Z_t^2\}$:
\[
A_i =
\begin{pmatrix}
  \alpha_1 Z_{t-1}^2 + \beta_1 & \beta_2 & \cdots &
  \beta_{q-1} & \beta_q & \alpha_2 & \alpha_3 &
  \cdots & \alpha_{p-1} & \alpha_p\\
  1 & 0 & \cdots & 
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots & 
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  1 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  Z_{t-1}^2 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  0 & 0 & 1 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots &
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\    
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 1 & 0 \\    
\end{pmatrix}
\]
While GARCH models have been very successful for modelling financial
time series, it does have its drawbacks. For example, the tail
index is very sensitive to the model parameters $\{\alpha_i\}_{i=1}^p$ and
$\{\beta_i\}_{i=1}^q$. In applications, these parameters need to be
estimated from a sample and are always uncertain to some extent. For
this reason, there can be a significant discrepancy between the tail
index obtained via \eqref{eq:grgbg} with the model parameters
substituted for their sample estimates and the Hill estimate
\eqref{eq:fhhty}.

@@

Later authors, e.g. Back \cite{back:1991} chose to model the log-price
process as a semimartingale, with increments of the martingale
component modelled as a product process:
\begin{eqnarray*}
  Y_t &=& Y_0 + A_t + M_t \\
  M_t - M_{t-1} &=& X_t = \sigma_t Z_t
\end{eqnarray*}
where $A_t$ is a finite variation process and $M_t$ is a martingale
and hence $Y_t$ is a semimartingale. $\sigma_t$ is non-negative and
$Z_t$ is an iid process with 0 mean and unit variance. A convenient
choice of $\sigma_t$ is
\begin{equation}
  \label{eq:rfht}
  \log\sigma_t = \sum_{l \in \mathbb Z} \psi_l \eta_{t-l}  
\end{equation}
where $-1 \leq \{\psi_l\}_{l \in \mathbb Z} \leq 1$
is a sequence of real numbers with at least one non-zero element,
$\{\eta_t\}$ is a iid sequence of random variables. By Kolmogorov's
3-series theorem, the infinite series above converges if and only if
$\sum_{l \in \mathbb Z} \psi_l^2 < \infty$.
In particular, if $\{\eta_i\}$ is normally distributed with mean 0,
$\{\sigma_t\}$ is stationary and its marginal distributions are
log-normal.

If, however, the sequence $\{Z_t\}$ or $\{\sigma_t\}$ is regularly
varying with index $\alpha$ and some additional conditions are
satisfied, $X_t$ is also regularly varying with the same
index. Specifically, if $Z_t$ is regularly varying and $\sigma_t$ has
a lighter tail, the conclusion follows from Breiman's lemma. See
\S\ref{sec:case1} of Jan\ss en et al. i.e. chapter \ref{ch:bernoulli}
of this thesis for more details.

If, instead, $\sigma_t$ has a heavier tail than $Z_t$, $X_{i,t}$ is
regularly varying assuming the following conditions:
\begin{itemize}
\item $e^\eta$ is regularly varying with index $\alpha$, where $\eta$
  is a generic element of the iid sequence $\{\eta_t\}$, and
  \begin{equation}
    \label{eq:dfhty}
    \E e^{\alpha \eta} = \infty, \quad\text{ or }
    {\P(\eta_1 + \eta_2 > x) \over \P(\eta_1 > x)} \to c \in (0, \infty)
  \end{equation}
    where $\eta, \eta_1$ and $\eta_2$ are generic elements of the
    iid sequence $\{\eta_t\}$.
\end{itemize}
Note the only finite limit for
${\P(\eta_1 + \eta_2 > x) \over \P(\eta_1 > x)}$ in $(0, \infty)$ is
$\E e^{\alpha \eta}$. Hence the two conditions of \eqref{eq:dfhty}
are mutually exclusive. In chapter \ref{ch:bernoulli} we will see that
these two conditions lead to rather different behavior for
eigenvalues of the sample covariance matrix of stochastic volatility
processes of the form \eqref{eq:rfht}.

There are a few advantages in using stochastic volatility models. For
one thing, they are among the simplest models allowing for conditional
heteroscedasticity; for another thing,

>>>>

\section{GARCH models}
Introduced by Bollerslev \cite{bollerslev:1986} in 1986,
{\em Generalized Autoregressive Conditional Heteroscedasticity}
(GARCH) models have been hugely popular for modelling volatility of
financial time series and have inspired numerous variants.
A GARCH($p, q$) model is a stationary process
$\{X_t\}_{t \in \mathbb Z}$satisfying
\begin{eqnarray*}
  X_t &=& \sigma_t Z_t \\
  \sigma_t^2 &=& \omega + \sum_{i=1}^p \alpha_i X_{t-i}^2 +
  \sum_{j=1}^q \beta_j \sigma_{t-j}^2
\end{eqnarray*}
where $\{X_t\}$ is a return series, e.g. stock returns, foreign exchange
rates, interest rates, etc; $\{Z_t\}$ is an iid, mean 0, unit-variance
sequence, $\sigma_t^2$ is the variance of the distribution of $X_t$
conditional on
$\{(X_i, \sigma_i^2)\}_{i=0}^{t-1}$; $\omega, \{\alpha_i\}_{i=1}^p,
\{\beta_i\}_{i=1}^q$
are non-negative parameters of the model. The GARCH($p,q$) recurrence
equation is of the form of \eqref{eq:rhjyu}. With appropriate
conditions, $\sigma_t^2$ is shown to be a positive Harris recurrent
Markov chain (cf. Bollerslev \cite{bollerslev:1986} and Buraczewski et al.
\cite{buraczewski:damek:mikosch:2016}), whose stationary distribution
has regularly varying tails. The tail index $\alpha$ is given by
\begin{equation}
  \label{eq:grgbg}
  \lim_{n \to \infty} {1 \over n}\log\E\|A_n \cdots A_1\|^\alpha = 0,
\end{equation}
where $\{A_i\}_{i \in \mathbb Z}$ are iid matrices whose entries are
functions of $\{\alpha_i\}_{i=1}^p$, $\{\beta_i\}_{i=1}^q$ and
$\{Z_t^2\}$:
\[
A_i =
\begin{pmatrix}
  \alpha_1 Z_{t-1}^2 + \beta_1 & \beta_2 & \cdots &
  \beta_{q-1} & \beta_q & \alpha_2 & \alpha_3 &
  \cdots & \alpha_{p-1} & \alpha_p\\
  1 & 0 & \cdots & 
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots & 
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  1 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  Z_{t-1}^2 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\
  0 & 0 & \cdots &
  0 & 0 & 1 & 0 & \cdots & 0 & 0 \\
  \vdots & \vdots & \ddots &
  \vdots & \vdots & \vdots & \vdots &
  \ddots & \vdots & \vdots \\
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 0 & 0 \\    
  0 & 0 & \cdots &
  0 & 0 & 0 & 0 & \cdots & 1 & 0 \\    
\end{pmatrix}
\]
While GARCH models have been very successful for modelling financial
time series, it does have its drawbacks. For example, the tail
index is very sensitive to the model parameters $\{\alpha_i\}_{i=1}^p$ and
$\{\beta_i\}_{i=1}^q$. In applications, these parameters need to be
estimated from a sample and are always uncertain to some extent. For
this reason, there can be a significant discrepancy between the tail
index obtained via \eqref{eq:grgbg} with the model parameters
substituted for their sample estimates and the Hill estimate
\eqref{eq:fhhty}.

@@

Later authors, e.g. Back \cite{back:1991} chose to model the log-price
process as a semimartingale, with increments of the martingale
component modelled as a product process:
\begin{eqnarray}
  Y_t &=& Y_0 + A_t + M_t \nonumber \\
  M_t - M_{t-1} &=& X_t = \sigma_t Z_t
  \label{eq:rtght}
\end{eqnarray}
where $A_t$ is a finite variation process and $M_t$ is a martingale
and hence $Y_t$ is a semimartingale. $\sigma_t$ is non-negative and
$Z_t$ is an iid process with 0 mean and unit variance. A convenient
choice of $\sigma_t$ is
\begin{equation}
  \label{eq:rfht}
  \log\sigma_t = \sum_{l \in \mathbb Z} \psi_l \eta_{t-l}  
\end{equation}
where $-1 \leq \{\psi_l\}_{l \in \mathbb Z} \leq 1$
is a sequence of real numbers with at least one non-zero element,
$\{\eta_t\}_{t \in \mathbb Z}$ is a iid sequence of random
variables. By Kolmogorov's 3-series theorem, the infinite series above
converges if and only if $\sum_{l \in \mathbb Z} \psi_l^2 < \infty$.
In particular, if $\{\eta_i\}_{i \in \mathbb Z}$ is normally
distributed with 0 mean, $\{\sigma_t\}_{t \in \mathbb Z}$ is
stationary and its marginal distributions are log-normal.

If, however, the sequence $\{Z_t\}_{t \in \mathbb Z}$ or
$\{\sigma_t\}_{t \in \mathbb Z}$ is regularly varying with index
$\alpha$ and some additional conditions are satisfied, $X_t$ is also
regularly varying with the same index. Specifically, if $Z_t$ is
regularly varying and $\sigma_t$ has a lighter tail, the conclusion
follows from Breiman's lemma. See \S\ref{sec:case1} of Jan\ss en et
al. i.e. chapter \ref{ch:bernoulli} of this thesis for more details.

If, instead, $\sigma_t$ has a heavier tail than $Z_t$, then $X_{i,t}$
is regularly varying assuming the following conditions:
\begin{itemize}
\item $e^\eta$ is regularly varying with index $\alpha$, where $\eta$
  is a generic element of the iid sequence
  $\{\eta_t\}_{t \in \mathbb Z}$, and
  \begin{equation}
    \label{eq:dfhty}
    \E e^{\alpha \eta} = \infty, \quad\text{ or }
    {\P(\eta_1 + \eta_2 > x) \over \P(\eta_1 > x)} \to c \in (0, \infty)
  \end{equation}
    where $\eta, \eta_1$ and $\eta_2$ are generic elements of the
    iid sequence $\{\eta_t\}_{t \in \mathbb Z}$.
\end{itemize}
Note the only finite limit for
${\P(\eta_1 + \eta_2 > x) \over \P(\eta_1 > x)}$ is
$\E e^{\alpha \eta}$. Hence the two conditions of \eqref{eq:dfhty}
are mutually exclusive. In chapter \ref{ch:bernoulli} we will see that
these two conditions lead to rather different behavior for
eigenvalues of the sample covariance matrix of stochastic volatility
processes of the form \eqref{eq:rfht}.

There are a few advantages in using stochastic volatility models. For
one thing, they are among the simplest models allowing for conditional
heteroscedasticity (cf. Andersen et
al. \cite{andersen:davis:kreiss:mikosch:2009}); for another thing,
they greatly improve the accuracy of predicting future volatilities
by taking advantage of high frequency data. Specifically, for the
semimartingale $Y_t$ we have
\begin{equation}
  \label{eq:jieg}
  [Y_{\delta}]_t
  =
  \sum_{j=1}^{\floor{t/\delta}}
  (Y_{j \delta} - Y_{(j-1)\delta})^2
  \overset{P}{\to} [Y]_t,
  \quad
  \delta \to 0.
\end{equation}
The object $[Y_\delta]_t$ is called the realized quadratic variation
in time series literature.
The convergence in probability follows directly from the definition of
quadratic variation.
%% When $\E Z_t = 0$, $[Y]_t = [M]_t$ 
Furthermore, Jacod \cite{jacod:1994} and
Barndorff-Nielsen and Shephard \cite{barndorff:shephard:2002} proved a
central limit theorem:
\[
  {[Y_\delta]_t - [Y]_t \over \sqrt{2 \delta \int_0^t \sigma_s^2 ds}}
  \overset{d}{\to} N(0, 1),
  \quad \delta \to 0
\]
Moreover, the structure $X_t = M_{t} - M_{t-1} = \sigma_t Z_t$ gives
$[Y_\delta]_t \overset{P}{\to} \int_0^t \sigma_s^2 ds$. Meanwhile,
It\^o's formula gives, for semimartingale $Y_t$,
\begin{eqnarray}
  Y_t^2 &=& [Y]_t + 2 \int_0^t Y_s d Y_s \nonumber \\
  &=& [Y]_t + 2 \int_0^t Y_s dA_s + 2 \int_0^t Y_s dM_s \nonumber \\
  \E Y_t^2 &=& \E [Y]_t + 2 \E \left( \int_0^t Y_s dA_s \right) \nonumber\\
  &\approx& \E [Y]_t, \quad \text{when $t$ is small} \label{eq:hytre}
\end{eqnarray}
\eqref{eq:hytre} and \eqref{eq:jieg} shows $\E Y_t^2$, or in other
words, the forecast of future squared return can be obtained as
$\E [Y_\delta]_t$, i.e. the forecast of future realized quadratic
variation. Since past realized quadratic variations are observable,
forcast of future $[Y_\delta]_t$ can be obtained using e.g. an ARFIMA
model.
